A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-07-10 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: reviewed fraction capacity

TODO: intro

Given that mixing coefficients have a more general applicability than probability of outcomes, let us concentrate on the following problem: given an ensemble $\ens \in \Ens$ and a (Borel) set of ensembles $A \subseteq \Ens$, what fraction of $\ens$ can be constructed by a mixture of $A$? How much of $\ens$ can be explained as coming from preparations corresponding to $A$? For example, let $\Ens$ be the space of all probability distributions for a six-sided die. Let $\ens_{123456}$ be the uniform distribution over all outcomes. Let $A_1 = \{\ens_1\}$ where $\ens_1$ represents outcome one with 100\% probability. Since we can write $\ens_{123456} = \frac{1}{6} \ens_{1} + \frac{5}{6} \ens_{23456}$, where $\ens_{23456}$ represents a uniform distributions over the five outcomes, $1/6$ of $\ens$ can be constructed from $A$, but no more. Similarly, if $A_{12} = \{\ens_{1},\ens_{2}\}$, we can write $\ens = \frac{1}{3} \left(\frac{1}{2} \ens_1 + \frac{1}{2} \ens_2 \right)  + \frac{2}{3} \ens_{3456}$, so $1/3$ of $\ens$ can be constructed from $A$, but no more. Note that, given the uniform distribution, $1/6$ is exactly the probably for the event $A_1$ and $1/3$ the probability for event $A_{12}$. This gives the basic insight for our definitions.

@@

We now extend this idea from a single ensemble $\ens[a] \in \Ens$ to a Borel set of ensembles $A \subset \Ens$. Given that the $\hull(A)$ is the set of all possible convex combinations (i.e. mixtures),\footnote{Since we work with convex sets within a topology, the hull is both the closed under the topology and convex combinations. That is, $\hull(A)$ is the smallest closed and convex set that contains $A$} we define the \textbf{fraction capacity} of $A$ for $\ens$ to be
\begin{equation}
	\frcap_{\ens}(A) = \sup(\fraction_{\ens}(\hull(A))\cup\{0\}).
\end{equation}
This returns the biggest fraction of $\ens$ that can be achieved with a mixture of elements of $A$.\footnote{The name fraction capacity is chosen to both signify the ability of the set $A$ to contain $\ens$ (i.e. fraction capacity of one means $\ens$ is within the convex combinations of $A$) and to indicate that it will be a non-additive monotonic measure, which are called ``capacities'' in some literature. TODO \href{ https://link.springer.com/book/10.1007/978-3-319-03155-2}{cite} }

One can then show that the fraction capacity $\frcap_{\ens} : \Sigma_{\Ens} \to [0,1]$ is a set function that satisfies the following:
\begin{enumerate}
	\item non-negative and unit bounded - $\frcap_{\ens}(A) \in [0,1]$
	\item monotone - $A \subseteq B \implies \frcap_{\ens}(A) \leq \frcap_{\ens}(B)$
	\item sub-additive - $\frcap_{\ens}(A \cup B) \leq \frcap_{\ens}(A) + \frcap_{\ens}(B)$
	\item continuous from below and above - $\frcap_{\ens}(\lim\limits_{i \to \infty} A_i) = \lim\limits_{i \to \infty} \frcap_{\ens}(A_i)$ for any increasing or decreasing sequence $\{A_i\}$.
\end{enumerate}

The first property readily comes from the range of the fraction. For the second, note that $\hull$ and $\sup$ are both monotone. For sub-additivity, note that the fraction of two components can at most sum during mixing. For the last, increasing and decreasing sequences strictly add or remove possible mixtures, therefore they will lead to increasing or decreasing sequences of real numbers bounded between zero and one. The limit of these sequences must agree with the fraction capacity of the limit. If not for the lack of additivity over disjoint sets, the fraction capacity would be a probability measure.

Our setup is very similar to what happens in Choquet theory. The fraction capacity, in fact, is not just defined on the extreme points, but it is defined on the whole space of ensembles. In the classical discrete case, for example, if $A \subset \Omega \subset \Ens$ is an event, then $\frcap_{\ens}(A)$ will return the probability of event $A$; if $P \subset \Ens$ represents all the Poisson distributions with integer coefficient, $frcap_{\ens}(P)$ will return how much of $\ens$ can be expressed is a mixture of those distributions. Similarly, the probability measures in Choquet theory are, in general, defined on the whole ensemble space, and then one finds those particular ones that have support on the extreme points.

We can establish a direct connection between the fraction capacity and Choquet theory, because the ability to express an ensemble $\ens$ as the convex combination of $A$ is precisely the ability to find a measure $\mu$ that has support within $A$. That is, $\ens = \sum_{i=1}^{\infty} p_i \ens[a]_i$ if an only if the probability measure defined by $\mu(\{\ens[a]_i\}) = p_i$ represents $\ens$.\footnote{TODO: add proof somewhere} Consequently, if $M_{\ens}$ is the set of all probability measures that represent $\ens$ as defined in Choquet theory, then $\frcap_{\ens}(A) = \sup(\{\mu(A) \, | \, \mu \in M_{\ens}\})$.\footnote{TODO: prove somewhere}

As we saw, the classical case corresponds to a simplex in Choquet theory. This means that each element of the ensemble space is represented by a unique probability measure over the extreme points. In that case, the fraction capacity extended on the extreme points will also be additive. In the non-classical case, the closure of the ensemble space will not be a simplex, and there will be multiple representations, which will lead to a non-additive measure. Suppose, in fact, that $\mu$ and $\nu$ are two distinct measures that represent the same ensemble $\ens$. Then there will be a set $A$ such such that $\mu(A) \neq \nu(A)$. Suppose, without loss of generality, that $\mu(A) > \nu(A)$ We will also have $\mu(A^{\complement}) = 1- \mu(A) < 1 - \nu(A) = \nu(A^{\complement})$. Since the fraction capacity for a set cannot be less than the measure on that set, we have $\frcap_{\ens}(A \cup A^{\complement})=\frcap_{\ens}(X) = 1 = \mu(A) + \mu(A^{\complement}) < \mu(A) + \nu(A^{\complement}) \leq \frcap_{\ens}(A) + \frcap_{\ens}(A^{\complement})$. That is, the fraction capacity is additive over disjoint sets of extreme points in and only in the classical case. In other words, the fraction capacity reduces to a classical probability distribution over pure states in the classical case.

>>>>

\textbf{Probability of outcomes vs fraction of components.} From the previous discussion, we can see that the problem with using probability in quantum mechanics is that not all pure states represent mutually exclusive events, and it is only over mutually exclusive events that classical probability can be defined. Mutually exclusive events correspond to ensemble are orthogonal, as each instance of one ensemble cannot be produced by the other ensemble. That is why, in quantum mechanics, probability can be defined only once a context, an orthogonal basis, is chosen. Therefore, at a conceptual level, probabilities can only give us a patchwork representation of ensembles. Some author embrace this view, by characterizing states as the collection of all probability distributions of all possible results, as it is generally the case in GPTs. [CITATION NEEDED] However, there is a different route.

Note that the linearity in our axioms is in terms of mixing coefficients, not probability of outcomes. That is, regardless of whether we can characterize experiment outcomes with probability distributions, preparation procedures will always allow us to create statistical mixture. We can create statistical mixtures regardless of whether the states we are mixing represent mutually exclusive outcomes or not. That is, instead of focusing on the likelihood of an outcome, we will focus on how much of a particular ensemble can be understood as the mixture of some other ensembles. Instead of focusing on what an ensemble may cause, we focus on how an ensemble can be produced. It turns out, this second approach generalizes in a natural way.

\textbf{Fraction capacity.} Let us concentrate on the following problem: given an ensemble $\ens \in \Ens$ and a (Borel) set of ensembles $A \subseteq \Ens$, what fraction of $\ens$ can be constructed by a mixture of $A$? How much of $\ens$ can be explained as coming from preparations corresponding to $A$? For example, let $\Ens$ be the space of all probability distributions for a six-sided die. Let $\ens_{123456}$ be the uniform distribution over all outcomes. Let $A_1 = \{\ens_1\}$ where $\ens_1$ represents outcome one with 100\% probability. Since we can write $\ens_{123456} = \frac{1}{6} \ens_{1} + \frac{5}{6} \ens_{23456}$, where $\ens_{23456}$ represents a uniform distributions over the five outcomes, $1/6$ of $\ens$ can be constructed from $A$, but no more. Similarly, if $A_{12} = \{\ens_{1},\ens_{2}\}$, we can write $\ens = \frac{1}{3} \left(\frac{1}{2} \ens_1 + \frac{1}{2} \ens_2 \right)  + \frac{2}{3} \ens_{3456}$, so $1/3$ of $\ens$ can be constructed from $A$, but no more. Note that, given the uniform distribution, $1/6$ is exactly the probably for the event $A_1$ and $1/3$ the probability for event $A_{12}$. This gives the basic insight for our definitions.

@@

We now extend this idea from a single ensemble $\ens[a] \in \Ens$ to a Borel set of ensembles $A \subset \Ens$. Given that the $\hull(A)$ is the set of all possible convex combinations (i.e. mixtures),\footnote{Since we work with convex sets within a topology, the hull is both the closed under the topology and convex combinations. That is, $\hull(A)$ is the smallest closed and convex set that contains $A$} we define the \textbf{fraction capacity} of $A$ for $\ens$ to be
\begin{equation}
	\frcap_{\ens}(A) = \sup(\fraction_{\ens}(\hull(A))\cup\{0\}).
\end{equation}
This returns the biggest fraction of $\ens$ that can be achieved with a mixture of elements of $A$.

\textbf{Non-additivity of the fraction capacity.} One can then show that the fraction capacity $\frcap_{\ens} : \Sigma_{\Ens} \to [0,1]$ is a set function that satisfies the following:
\begin{enumerate}
	\item non-negative and unit bounded - $\frcap_{\ens}(A) \in [0,1]$
	\item monotone - $A \subseteq B \implies \frcap_{\ens}(A) \leq \frcap_{\ens}(B)$
	\item sub-additive - $\frcap_{\ens}(A \cup B) \leq \frcap_{\ens}(A) + \frcap_{\ens}(B)$
	\item continuous from below and above - $\frcap_{\ens}(\lim\limits_{i \to \infty} A_i) = \lim\limits_{i \to \infty} \frcap_{\ens}(A_i)$ for any increasing or decreasing sequence $\{A_i\}$.
\end{enumerate}

The first property readily comes from the range of the fraction. For the second, note that $\hull$ and $\sup$ are both monotone. For sub-additivity, note that the fraction of two components can at most sum during mixing. For the last, increasing and decreasing sequences strictly add or remove possible mixtures, therefore they will lead to increasing or decreasing sequences of real numbers bounded between zero and one. The limit of these sequences must agree with the fraction capacity of the limit.

Note that, if not for the lack of additivity over disjoint sets, the fraction capacity would be a probability measure. It is therefore a non-additive probability measure. The name fraction capacity is chosen to both signify the ability of the set $A$ to contain $\ens$ (i.e. fraction capacity of one means $\ens$ is within the convex combinations of $A$) and to indicate that it will be a non-additive monotonic measure, which are called ``capacities'' in some literature.\footnote{ \href{ https://link.springer.com/book/10.1007/978-3-319-03155-2}{cite} } We will see how our approach fits  in that literature in a later section.

\textbf{Compare to Choquet theory.} There is a straight forward link between our non-additive measure and Choquet theory. Recall that Choquet theory start by defining measure on the whole ensemble space before finding those constrained to the extreme points. Our setup is similar. The fraction capacity is defined on the whole ensemble space. In the classical discrete case, for example, if $A \subset \Omega \subset \Ens$ is an event, then $\frcap_{\ens}(A)$ will return the probability of event $A$; if $P \subset \Ens$ represents all the Poisson distributions with integer coefficient, $\frcap_{\ens}(P)$ will return how much of $\ens$ can be expressed as a mixture of Poisson distributions. In case that extreme points exist, or we can extend the ensemble space with some limit procedure, we can also define the fraction capacity on the extreme points.

We can establish a direct connection between the fraction capacity and Choquet theory, because the ability to express an ensemble $\ens$ as the convex combination of $A$ is precisely the ability to find a measure $\mu$ that has support within $A$. That is, $\ens = \sum_{i=1}^{\infty} p_i \ens[a]_i$ if an only if the probability measure defined by $\mu(\{\ens[a]_i\}) = p_i$ represents $\ens$.\footnote{TODO: add proof somewhere} Consequently, if $M_{\ens}$ is the set of all probability measures that represent $\ens$ as defined in Choquet theory, then $\frcap_{\ens}(A) = \sup(\{\mu(A) \, | \, \mu \in M_{\ens}\})$.\footnote{TODO: prove somewhere} The fraction capacity is simply the supremum on each set over all possible measures.

As we saw, the classical case corresponds to a simplex in Choquet theory. This means that each element of the ensemble space is represented by a unique probability measure over the extreme points. In that case, the fraction capacity restricted to the extreme points will also be additive. In the non-classical case, the closure of the ensemble space will not be a simplex, and there will be multiple representations, which will lead to a non-additive measure. Suppose, in fact, that $\mu$ and $\nu$ are two distinct measures that represent the same ensemble $\ens$. Then there will be a set $A$ such such that $\mu(A) \neq \nu(A)$. Suppose, without loss of generality, that $\mu(A) > \nu(A)$ We will also have $\mu(A^{\complement}) = 1- \mu(A) < 1 - \nu(A) = \nu(A^{\complement})$. Since the fraction capacity for a set cannot be less than the measure on that set, we have $\frcap_{\ens}(A \cup A^{\complement})=\frcap_{\ens}(X) = 1 = \mu(A) + \mu(A^{\complement}) < \mu(A) + \nu(A^{\complement}) \leq \frcap_{\ens}(A) + \frcap_{\ens}(A^{\complement})$. That is, the fraction capacity is additive over disjoint sets of extreme points in and only in the classical case. In other words, the fraction capacity reduces to a classical probability distribution over pure states in the classical case.

\textbf{Uniqueness of the fraction capacity.} From a conceptual standpoint, the fraction capacity has a major advantage: it has a clear physical meaning and it is unique. Like in Choquet theory, and unlike for pseudo-probability, the domain over which the measure is defined is fully determined by the system, by the ensemble space. However, unlike Choquet, and like pseudo-probability once the space is fixed, there one unique fraction capacity for each ensemble. Moreover, the setup is more general than Choquet theory (i.e. the ensemble space need not be compact) and stems from necessary physical requirement. An in depth study on fraction capacities, then, is guaranteed to be applicable to all physical theories, including those yet to be formulated.
